The Unicorn Equation: Conceptual unification of ecology, evolution, and ecosystem function: TESTING

### 2024-12-20 tuesday_seminar.Rmd
Initialise all of the relevant files

---
title: "Foundations of ecological and evolutionary change"
author: "Brad Duthie and Victor Luque"
output:
  beamer_presentation:
    theme: "default"
    colortheme: "default"
    fonttheme: "default"
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \usepackage{hyperref}
- \usepackage{caption}
- \usepackage[normalem]{ulem}
- \definecolor{links}{HTML}{2A1B81}
- \hypersetup{colorlinks,linkcolor=,urlcolor=links}
colorlinks: true
linkcolor: blue
urlcolor: blue
---

@@

@@

## Complexity of the biological sciences

```{r, out.width = "90%", echo = FALSE, fig.cap = ""}
include_graphics("img/bee.jpg");
```

\footnotetext[1]{Image: John Severns (2006). \href{https://commons.wikimedia.org/wiki/File:European_honey_bee_extracts_nectar.jpg}{Public Domain}}

## Complexity of the biological sciences

```{r, out.width = "90%", echo = FALSE, fig.cap = ""}
include_graphics("img/food_web.png");
```

\footnotetext[1]{Darnis, G, et al. (2012). Current state and trends in Canadian Arctic marine ecosystems: II. Heterotrophic food web, pelagic-benthic coupling, and biodiversity. Climatic Change, 115, 179-205.}
\footnotetext[1]{Image: Darnis, G, et al. (2012). \href{https://commons.wikimedia.org/wiki/File:Arctic_Marine_Food_Web.png}{Public Domain}}

## Unification of evolution and genetics

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{center}
\textbf{Darwinian evolution}
\end{center}

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/ithink.png");
```

\end{column}
\begin{column}{0.5\textwidth}  
\begin{center}
\textbf{Mendelian genetics}
\end{center}

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/punnet_square_2.png");
```

\end{column}
\end{columns}

\footnotetext[1]{Image (left): \href{https://commons.wikimedia.org/wiki/File:Darwin_tree.png}{Public Domain}}
\footnotetext[2]{Image (right): Madeleine Price Ball (2007).  \href{https://commons.wikimedia.org/wiki/File:Punnett_square_mendel_flowers.svg}{Public Domain}}

## Unification of evolution and genetics

@@

$$p^{2} + 2pq + q^{2} = 1$$
\vspace{0.5mm}
\hrule
\vspace{1mm}

\begin{itemize}
\item No natural selection
\item No mutation
\item No migration (no gene flow)
\item Infinite population size
\item Mating is random
\item Non-overlapping generations
\end{itemize}

## General theory of community ecology

\begin{columns}
\begin{column}{0.5\textwidth}
\begin{center}
\textbf{Population genetics}
\vspace{2mm}
\hrule
\vspace{2mm}
\end{center}

\begin{itemize}
\item Selection
\item Mutation
\item Gene flow
\item Genetic drift
\end{itemize}

\end{column}
\begin{column}{0.5\textwidth}  
\begin{center}
\textbf{Community ecology}
\vspace{2mm}
\hrule
\vspace{2mm}
\end{center}

\begin{itemize}
\item Species selection
\item Disperal
\item Drift
\item Speciation
\end{itemize}

\end{column}
\end{columns}
\vspace{3mm}
\hrule

General community ecology theory focused on high-level processes$^{1}$

\footnotetext[1]{Vellend, M (2016). The theory of ecological communities. Monographs in population biology. Princeton University Press.}

## Eco-evolutionary models

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/fig1_b.png");
```

## Eco-evolutionary models

\begin{columns}
\begin{column}{0.5\textwidth}

Lion$^{1}$ (2018)
\hrule

\vspace{2mm}

\textbf{Ecology:}

$$\frac{dn_{i}}{dt} = r_{i}\left(\mathbf{E}\right)n_{i}$$

\vspace{4mm}
\textbf{Evolution:}

$$\frac{df_{i}}{dt} = f_{i}\left(r_{i} \left(\mathbf{E} \right) - \bar{r} \left(\mathbf{E} \right) \right)$$

\end{column}
\begin{column}{0.5\textwidth}

Patel et al.$^{2}$ (2018)
\hrule

\vspace{2mm}

\textbf{Ecology:}

$$\frac{dN_{i}}{dt} = N_{i}f_{i}\left(N, x\right)$$

\vspace{2mm}
\textbf{Evolution:}

$$\frac{dx_{j}}{dt} = \epsilon g_{j}\left(N, x\right)$$

\end{column}
\end{columns}

\footnotetext[1]{Lion, S (2018). Theoretical approaches in evolutionary ecology: environmental feedback as a unifying perspective. Am. Nat., 191:21-44}
\footnotetext[2]{Patel, S, Cortez, MH, \& Schreiber, SJ (2018). Partitioning the effects of eco-evolutionary feedbacks on community stability. The Am. Nat., 191:381-394.}

## Conceptual unification of ecology and evolution

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/fig1_b.png");
```

## Conceptual unification of ecology and evolution

## Unification of modern coexistence theory \& Price equation

Joint funding call between French Foundation for Research on Biodiversity and German Centre for Integrative Biodiversity Research (iDiv): CESAB-sDIV SYNERGY

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/unicop_team.jpg");
```

The UNICOP team in Montpellier (summer 2023)

@@

- Not making assumptions or predictions \pause
- Defining population change in an eco-evo context \pause
- Framework for models (meta-modelling) \pause
- A rough analogy,

$$y = \beta_{0} + \beta_{1}x$$

@@

\footnotetext[1]{Queller, DC (2017). Fundamental theorems of evolution. The American Naturalist, 189:345-353.}

## Generality of the Price equation

@@

**Evolution of non-biological systems**

- Archaeological artefacts$^{1}$
- Musical properties$^{2}$

@@

**Synthesis across disciplines**

- D'Alembert's principle of mechanics$^{3}$
- Information theory$^{4}$

\footnotetext[1]{Brantingham, PJ (2007). A unified evolutionary model of archaeological style and function based on the Price equation. American Antiquity 72:395–416.}
\footnotetext[2]{MacCallum, RM, et al. (2012). Evolution of music by public choice. PNAS 109:12081–12086.}
\footnotetext[3]{Frank, SA (2015). D'Alembert’s direct and inertial forces acting on populations: The price equation and the fundamental theorem of natural selection. Entropy 17:7087–7100.}
\footnotetext[3]{Frank, SA (2012). Natural selection. V. How to read the fundamental equations of evolutionary change in terms of information theory. Journal of Evolutionary Biology, 25:2377-2396.}

## Conceptual unification of ecology and Price equation

## Links between ecology and the Price equation

**Ecosystem function loss due to species loss$^{1}$**

$$\Delta T = \bar{z}\Delta s + Sp(w, z) + \sum_{i}w_{i}\Delta z_{i}$$

\normalsize\pause

**Community trait change: Physiology, evolution, ecology$^{2}$**

$$\Delta \bar{z} = E_{I}\left(E_{Ji} \left(\Delta z_{ij} \right) \right) + E_{I}\left(cov_{Ji} \left(w_{ij}, z'_{ij} \right) \right) + cov_{I}\left(w_{i}, z'_{i}\right)$$

\footnotetext[1]{Fox, JW (2006). Using the price equation to partition the effects of biodiversity loss on ecosystem function. Ecology, 87(11), 2687-2696.}
\footnotetext[2]{Collins, S \& Gardner, A (2009). Integrating physiological, ecological and evolutionary change: A Price equation approach. Ecology Letters, 12(8), 744-757.}

## Links between ecology and the Price equation

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/Page_Nowak.png");
```

- Figure 1 from Page \& Nowak$^{1}$
- Relies on \textit{relative} species frequenies

\footnotetext[1]{Page, KM \& Nowak, MA (2002). Unifying evolutionary dynamics. Journal of Theoretical Biology, 219(1), 93–98.}

## Reconciling evolutionary change and population growth {.t}

- Start with Price equation and add ecology?
- Start elsewhere and derive the Price equation?

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/lost_in_woods.jpg");
```

\footnotetext[1]{Image: Eneas De Troya (2006).  \href{https://commons.wikimedia.org/wiki/File:Lost_in_a_Virtual_Wood_(177504871).jpg}{CC BY 2.0}}

## The fundamental equation of ecology

@@

- Evolution: Price Equation
- Ecology: Births and deaths

@@

## Reconciling foundations of evolution and ecology

@@

## The unifying equation

- No systematic approach
- Individual birth, death, characteristics
- Deductive exploration

\vspace{5mm}
\hrule
\pause
\vspace{5mm}

@@

## A formal definition of eco-evolutionary change {.t}

@@

- No simplifying assumptions \pause
- Complete and exact (closed populations) \pause
- Ecology: $\beta_{i} - \delta_{i} + 1$ \pause
- Evolution: $z_{i} + \Delta z_{i}$ \pause
- Ecology required for selection or drift

## Confirmations from a formal definition {.t}

@@

- **Fitness:** $\bar{w} = \lambda = \beta - \delta + 1$ \pause
- Derives assumed $N_{t+1} = N_{t}\lambda = N_{t}\bar{w}$ \pause
- Lifetime fitness is still $\beta_{i}$ \pause
- Starting point for further unification \pause
- Build upon for community ecology

## An elegance in two fits {.t}

@@

\vspace{2mm}
\hrule

@@

## An elegance in two fits {.t}

@@

## Foundations of ecological and evolutionary change {.t}

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/foundations.png");
```

\footnotetext[1]{Duthie, AB, \& Luque, VJ (2024). Foundations of ecological and evolutionary change. arXiv preprint arXiv:2409.10766.}

>>>>

---
title: "The Unicorn Equation: Conceptual unification of ecology, evolution, and ecosystem function: TESTING"
output:
  beamer_presentation:
    theme: "default"
    colortheme: "default"
    fonttheme: "default"
  ioslides_presentation: default
  slidy_presentation: default
header-includes:
- \usepackage{hyperref}
- \usepackage{caption}
- \usepackage[normalem]{ulem}
- \definecolor{links}{HTML}{2A1B81}
- \hypersetup{colorlinks,linkcolor=,urlcolor=links}
colorlinks: true
linkcolor: blue
urlcolor: blue
---

@@

## BES Tuesday seminars this autumn

**TEG: LT W1 from 13:00-14:00** (until 26 NOV)

\large

| Date    |  Speaker           | | Date    | Speaker            |
|---------|--------------------|-|---------|--------------------|
| 24 SEP  | Heather Anderson   | | 05 NOV  | Reuben Nowell      |
| 01 OCT  | **Open**           | | 12 NOV  | Ruth Wright        |
| 08 OCT  | Matt Tinsley       | | 19 NOV  | Zarah Pattison     |
| 15 OCT  | Carmen Carmona     | | 26 NOV  | Kirsty Park        |
| 22 OCT  | Mahtab Yaghouti    | | 03 DEC  | Emily Waddell      |
| 29 OCT  | Ponsarut Boonchuay | | 10 DEC  | Colin Bull         |

@@

## Big questions in biology

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/cvs.png");
```

## Big questions in biology

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/iowa_state.png");
```

## Call for proposals: joint sDiv-CESAB

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/synergy_call.png");
```

## Call for proposals: joint sDiv-CESAB

## The UNICOP project

@@

Conceptual unification of modern coexistence theory and the Price equation

@@

- Framework for making models (meta-modelling) \pause
- Not making assumptions or predictions \pause
- Not the kind of thing we test, e.g.,

$$y_{i} = \beta_{0} + \beta_{1}x_{i} + \epsilon_{i}$$

@@

\footnotetext[1]{Queller, D. C. (2017). Fundamental theorems of evolution. The American Naturalist, 189:345-353.}

## My goal: starting point is unclear

- Figure out how to unify the Price equation with ecological change \pause
- Start with the Price equation and add ecology? \pause
- Start somewhere else and derive Price? \pause
- Is this even possible? \pause
- If it's possible, it will look obvious in hindsight

## The UNICOP team

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/team.jpg");
```

## Leipzig in September 2022

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/leipzig_2022.jpg");
```

## Leipzig in September 2022

- Confusion about the question
- Separated project aims
- An attempt at an answer: \pause

@@

$$iN + \Delta\bar{z} = i\left(1 + \bar{w} - \bar{\delta} \right)N + \frac{1}{\bar{w}}Cov\left(z, w\right) + \frac{1}{\bar{w}} E\left(w\Delta z \right).$$

@@

- "It's not *wrong*" - S$\'{e}$b \pause
- "Brad, that's cheating" -Th$\'{e}$o

## Montpellier in June 2023: Not much progress

```{r, out.width = "85%", echo = FALSE, fig.cap = ""}
include_graphics("img/montpellier.jpg");
```

## On the backburner until January 2024

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/leipzig_more_snow.jpg");
```

## On the backburner until January 2024

```{r, out.width = "160%", echo = FALSE, fig.cap = ""}
include_graphics("img/leipzig_snow.jpg");
```

## Working day and night back and forth

```{r, out.width = "100%", echo = FALSE, fig.cap = ""}
include_graphics("img/leipzig_2024.jpg");
```

## Finally a way forward, and fundamental ecology

@@

- Evolution: Price Equation
- Ecology: Exponential growth

@@

## Finally a way forward, and fundamental ecology

@@

## Nearly there: the unicorn equation

@@

## Who cares? {.t}

@@

- No simplifying assumptions \pause
- Complete and exact for all closed populations \pause
- Ecology: $\beta_{i} - \delta_{i} + 1$ \pause
- Evolution: $z_{i} + \Delta z_{i}$

## Who cares? {.t}

@@

- **Fitness:** $\bar{w} = \lambda = 1 + \beta + \delta$ \pause
- Derives assumed $N_{t+1} = N_{t}\lambda = N_{t}\bar{w}$ \pause
- Lifetime fitness is still $\beta_{i}$ \pause
- Starting point for further unification \pause
- Ecology hiding in population genetics

## The most elegant part {.t}

@@

@@

## The most elegant part {.t}

@@
